---
title: 'Milestone #5'
author: "Saira Mayet, Jessica Pak, Andrew Nguyen"
date: "11/24/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=40), tidy = T)


library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
library("sf")
library("sp")
library("spData")
library("dplyr")
library("ggplot2")
library("USAboundaries")
library("tmap")
library("shinyjs")

```

```{r import statement, echo=FALSE}
# from Milestone #2
file_path_vax <- 
  "https://data.chhs.ca.gov/dataset/ead44d40-fd63-4f9f-950a-3b0111074de8/resource/ec32eece-7474-4488-87f0-6e91cb577458/download/covid19vaccinesbyzipcode_test.csv"
file_path_county <- 
  "https://raw.githubusercontent.com/Averysaurus/reproducable_examples-/main/ca_county_demographics.csv"
vax_temp <- read.csv(file_path_vax)
county_temp <- read.csv(file_path_county)
# subset data to only include variables of interest
vax <- vax_temp %>% 
  select(-c("vaccine_equity_metric_quartile", 
            "local_health_jurisdiction", 
            "vem_source",
            "age12_plus_population",
            11:14))
county <- county_temp %>% 
  select(c("name", 
           "med_age"))
#get latest date and convert into character for use in plot captions
date <- vax %>%
  mutate(date = as.Date.character(as_of_date)) %>%
  select(- as_of_date)

current_date <- as.Date(max(date$date))
current_date_chr <- as.character(current_date)

caption <- paste("Source: California Department of Health, current to", current_date_chr)
```
**Project Statement** 

We are monitoring COVID-19 vaccination rates among counties in California and 
in relation to age. Specifically, we are interested in whether there is any 
correlation between median age and vaccinated person prevalence at the county 
level. Utilizing two datasets, the California Census Data and the COVID-19 
Vaccine Progress Dashboard, we intend on exploring, analyzing, and visualizing
vaccination rates at the county level and the relationship between age and 
vaccination rate.

**Methods** 

Of the two datasets of interest, the first describes COVID-19 vaccination 
administration across the state of California. This is sourced from the 
California Open Data Portal for the California Department of Public Health. 
Relevant fields include date, zip code, county, and raw counts of vaccination 
status, among other population information. The dataset starts from January 5th,
2021 and is continually updated to reflect new changes in the data, most 
recently updated on November 23rd, 2021. 

The second dataset describes demographics (e.g. population, race/ethnicity, age,
household size, etc.) for each California county, most recently updated in 
September 2021. This dataset was rehosted on Avery Richard’s GitHub, and is 
sourced from Census data.

Partial subsetting of data was initially completed to keep the following 
variables in each dataset: 


***COVID-19 Vaccination Dashboard***: Date, Zip Code, County, Population 5+, Number of
Persons Fully Vaccinated, Number of Persons Partially Vaccinated 

***County Census Data***: Name, Median Age 

\newpage 
**New Variables**

After subsetting the data from both sources, the data were cleaned and new 
variables were created. 


Data Cleaning

1. Mean imputation: county-level means of fully and partially vaccinated that 
will be used to replace NA values in dataset
2. Percent eligible population partially vaccinated = # of persons partially 
vaccinated / population 5+ (at county level)
3. Percent eligible population fully vaccination = # of persons fully vaccinated 
/ population 5+ (at county level)
4. Merging relational data: county demographic dataset with vaccine 
administration dataset using key variable "county". 


```{r subset latest data, echo=FALSE}
vax_latest <- vax %>%
  mutate(date = as_date(as_of_date)) %>%
  filter(date == max(date)) %>%
  select( - as_of_date)
```

```{r mean imputation,  echo=FALSE}
#aggregate county-level counts + %s for vaccination
vax_temp <- vax_latest %>%
  group_by(county) %>%
  mutate(county_partial = 
           sum(persons_partially_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm =T),
         county_fully = 
           sum(persons_fully_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm = T)) %>%
  ungroup()

#mean imputation: for ZIPs that have eligible population counts but lack vaccination numbers
vax_temp <- vax_temp %>%
  mutate(persons_partially_vaccinated_2 = 
           ifelse(is.na(persons_partially_vaccinated), 
                  age5_plus_population*county_partial,
                  persons_partially_vaccinated),
         persons_fully_vaccinated_2 = 
           ifelse(is.na(persons_fully_vaccinated),
                  age5_plus_population*county_fully, 
                  persons_fully_vaccinated))

#make second county-level aggregate with imputed data + original data
vax_temp <- vax_temp %>%
  group_by(county) %>%
  mutate(county_partial_2 = 
           sum(persons_partially_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm =T),
         county_fully_2 = 
           sum(persons_fully_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm = T),
         county_eligible_pop = 
           sum(age5_plus_population, na.rm = T)) %>%
  ungroup()

#create new dataset with only one entry per county
vax_aggregate <- vax_temp %>%
  distinct(county, .keep_all = T) %>%
  select(county, 
         county_partial_2, 
         county_fully_2, 
         county_eligible_pop) %>%
  filter(county_partial_2 != 0)
```

```{r merging datasets, echo=FALSE}
#Merging both county and vax_latest dataset
county <- county %>%
  rename(county = name)
data <- left_join(vax_aggregate, county, by="county")
```

```{r table data setup,echo=FALSE}
# final clean data
clean_data <- data %>% 
  rename(County = county,
         Partially_Vax_Rate = county_partial_2,
         Fully_Vax_Rate = county_fully_2,
         Eligible_Population = county_eligible_pop,
         Median_Age = med_age) %>% 
  mutate(Partially_Vax_Rate = round((Partially_Vax_Rate*100), digits =2),
         Fully_Vax_Rate = round((Fully_Vax_Rate*100), digits = 2),
         Eligible_Population = round(Eligible_Population, digits = 0),
         Median_Age = round(Median_Age, digits = 0)) %>% 
  arrange(County)
```
\newpage 

**Visualizations** 

The Table, "Vaccination Rates for Caliornia Counties," displays the partial vaccinated rate, fully vaccinated rate, total eligible population, and median age of each county in California. 
The summarized average rates, total eligible population, and median age for the 
entire state of California is calculated and presented at the top of the table.
```{r kable; table 1, echo=FALSE}
# table with descriptive statistics (averages calculated for CA as a whole) and 4 data elements
kable(clean_data, 
      booktabs = T, 
      col.names = c("County Name", 
                    "Partially Vaccinated Rate", 
                    "Fully Vaccinated Rate", 
                    "Total Eligible Population", 
                    "Median Age"), 
      align = "c", 
      caption = "Vaccination Rates for California Counties",
      format.args=list(big.mark=",")) %>% 
      add_header_above(c("California"= 1,
      "6.97"=1,
      "62.8"=1,
      "33,330,578"=1,
      "38"=1),
      bold = T) %>%
      kable_styling(latex_options = c("striped", "scale_down")) 
```

\newpage 
The scatterplot, "Percentage of Fully Vaccinated Californians by Median Age, 2021," shows the rate of fully vaccinated individuals stratified by median age of California residents, as measured from California Department of Public Health’s latest online data repository for COVID-19. A trend line was fitted to follow the points on the plot, and a general trend of increasing fully vaccinated rates can be seen near the median age of 36-38.

```{r age vs. full vaccination rate plot, echo=FALSE}
# Scatterplot of Median Age vs Fully Vaccinated Rate

ggplot(data = clean_data, aes(x= Median_Age,y= Fully_Vax_Rate)) +
  geom_point() +
  geom_smooth(col = "navy blue", size = 0.5) + 
  scale_y_continuous(limits=c(0, 100)) +
  labs(x= "Median age", y= "Rate(s) of full vaccination", title= "Percentage of Fully Vaccinated Californians by Median Age, 2021", caption="Data Source:
       California Department of Public Health") + theme_classic()
```
\newpage 

The map shows the rates of full vaccination, by county, symbolized by a color ramp (from low to high rates of vaccination going from yellow to dark red). A color ramp was made with classes of 10 percent each. A second map shows the median age of each country, symbolized by a purple color ramp (younger to older median age going from light purple to darker purple). The five-step color ramp ranges from 30 to 55, with each step covering five years of age. 

```{r Map, echo=FALSE}

#get CA boundary with high definition
#ca <- USAboundaries::us_states(resolution = "high", states = "CA")

#CA county boundary
ca_co <- USAboundaries::us_counties(resolution = "high", states = "CA") %>%
  select(- state_name)

#Join dataset with sf polygon
clean_data_rename <- clean_data %>%
  rename(name = County)

ca_covid <- dplyr::left_join(clean_data_rename, ca_co, by = "name")

#Make joined dataset into sf object
ca_covid <- st_as_sf(ca_covid, sf_column_name = "geometry")

#Create first map
legend_title = " "
tm_shape(ca_covid) + 
  tm_fill(col = "Fully_Vax_Rate", 
          title = legend_title,
          labels = c('20 - 30%','30 - 40%','40 - 50%','50 - 60%', '60 - 70%', '70 - 80%', '80 - 90%')) +
  tm_compass(type = "8star", 
             position = c("left", "bottom"),
             size = 2) +
  tm_layout(scale = 1, 
            title = "Percent of population fully vaccinated in California, by county",
            frame.lwd = 0,
            inner.margins = 0.1,
            legend.position = c("right", "top")) +
  tm_credits(caption, size = 0.5)

#Create second map
tm_shape(ca_covid) + 
  tm_fill(col = "Median_Age", 
          title = legend_title,
          palette = "Purples") +
  tm_compass(type = "8star", 
             position = c("left", "bottom"),
             size = 2) +
  tm_layout(scale = 1, 
            title = "Median age, by county",
            frame.lwd = 0,
            inner.margins = 0.1,
            legend.position = c("right", "top")) +
  tm_credits(caption, size = 0.5)
```

\newpage 

This last visualization features points plotted by month and percent of full vaccination in aggregate ZIP codes falling under four categories of the Vaccine Equity Metric (VEM) quartiles, which is based on demographic indicators of socioeconomic disadvantage and health vulnerability (where 1 is least disadvantaged/vulnerable and 4 and most disadvantaged/vulnerable). 
```{r}
#get data per day at ZIP code level
vax_longitudinal <- read.csv(file_path_vax) %>%
  group_by(county) %>%
  mutate(county_partial = 
           sum(persons_partially_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm =T),
         county_fully = 
           sum(persons_fully_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm = T)) %>%
  ungroup() %>%
  mutate(persons_partially_vaccinated_2 = 
           ifelse(is.na(persons_partially_vaccinated), 
                  age5_plus_population*county_partial,
                  persons_partially_vaccinated),
         persons_fully_vaccinated_2 = 
           ifelse(is.na(persons_fully_vaccinated),
                  age5_plus_population*county_fully, 
                  persons_fully_vaccinated)) %>%
   group_by(county) %>%
  mutate(county_partial_2 = 
           sum(persons_partially_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm =T),
         county_fully_2 = 
           sum(persons_fully_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm = T),
         county_eligible_pop = 
           sum(age5_plus_population, na.rm = T)) %>%
  ungroup()

#create aggregate dataset at VEM quartile level
vax_long_agg <- vax_longitudinal %>%
  group_by(as_of_date, vaccine_equity_metric_quartile) %>%
  summarize(percent_full_vem = 
              sum(persons_fully_vaccinated, na.rm = T)/
              sum(age5_plus_population, na.rm = T)) %>%
  ungroup() %>%
  filter(is.na(vaccine_equity_metric_quartile) == FALSE) %>%
  mutate(date = as.Date.character(as_of_date)) %>%
  select(- as_of_date)

#Create longitudinal scatterplot visualization
ggplot(data = vax_long_agg, aes(x = date, y = percent_full_vem)) +
  geom_point(stat = "identity", aes(col = vaccine_equity_metric_quartile)) +
  theme_classic() +
  labs(x = "Date", 
       y = "Percent fully vaccinated", 
       title = "Rates of full vaccination across Vaccine Equity Metric Quartiles", caption = caption) +
  guides(color=guide_legend(title="Vaccine Equity Metric Quartiles")) +
  scale_color_continuous(labels = 
                           c("1 - Least Disadvantaged", "2", "3", "4 - Most Disadvantaged")) +
  scale_y_continuous(labels = scales::percent)

```

\newpage
**Results** 

Our analyses and data visualization shows that the highest vaccination rates 
were seen among the age group between 35-40. The general trend seen for this age
group exhibited a full vaccination rate of approximately 70-73%, with rates seen 
above 75% as well. 

Outliers were seen with higher rates of vaccination, approximately 85% for a 
median age of 44. Outliers were also found for lower rates of vaccination of 
approximately 25% for the median age of 37. 

In terms of geography, higher vaccination rates were seen in counties such as 
San Francisco, Santa Clara, and Marin county with full vaccination rates at or 
above 80% and median ages of 38, 36, 44, respectively.  Trailing behind are the 
counties of  Alameda, Contra Costa, Imperial, and San Mateo with median ages of,
37, 38, 32, 39. 

In terms of demographic indicators, it can be seen that rates of vaccination follow, from highest to lowest, the order of vulnerability derived from the Vaccine Equity Metric (VEM), from least to most -- suggesting areas of improvement for equitable vaccine delivery. The aggregate of the least disadvantaged areas (VEM group 4) sees vaccination rates of  77%, whereas the most disadvantaged areas (VEM group 1) trails behind at 59%. 

**Discussion**

According to our results, the minimum median age of persons receiving COVID-19
Vaccination in the state of California is 30, for Merced county. The maximum age 
was found to be 51 in Sierra county. The overall median for the state of 
California is 38. A wide range of ages was seen in the overall analysis. 

