---
title: 'PHW251 Team Project: Milestone #2'
subtitle: 'Scenario Two: COVID Vaccination Progress'
author: "Saira Mayet, Jessica Pak, Andrew Nguyen"
date: "9/28/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Description of dataset**

- What is the data source?


    There are two datasets of interest: one describing COVID-19 vaccine administration across the state of california, sourced from the California Open Data Portal ("cov_vax_admin.csv"). Fields include date, ZIP code, county, and raw counts of vaccination status, amongst other population information. The dataset spans January 5th, 2021, to September 21st, 2021. 
    
    The other describes demographics (e.g. population, race/ethnicity, age, household size, etc.) for each California county, updated to 2012. This dataset was rehosted on Avery Richards' GitHub, and is sourced from Census Data.
    
    
- How does the dataset relate to the group statement and question?

    
    Problem statement: We are monitoring the state level COVID-19 vaccination rates among counties in California and in relation to age.  
    Question: Is there any correlation between median age and vaccinated person prevalence on the county level?

    The group statement and question relates to exploring, analyzing, and visualizing vaccination rates at county level and to explore if there is a correlation between age and vaccination rate. These two described datasets have necessary fields to support these analyses by including vaccination information at the ZIP level and county demographic data. 

**Load libraries**
```{r load libraries}
library(tidyverse)
library(lubridate)
```

**Import Statement**

```{r import statement}
file_path_vax <- "https://data.chhs.ca.gov/dataset/ead44d40-fd63-4f9f-950a-3b0111074de8/resource/ec32eece-7474-4488-87f0-6e91cb577458/download/covid19vaccinesbyzipcode_test.csv"
file_path_county <- "https://raw.githubusercontent.com/Averysaurus/reproducable_examples-/main/ca_county_demographics.csv"

vax_temp <- read.csv(file_path_vax)
county_temp <- read.csv(file_path_county)

str(vax_temp)
vax <- vax_temp %>% select(-c("local_health_jurisdiction", "vem_source", 10:13))
county <- county_temp %>% select(c("name", "med_age"))

head(vax)
head(county)
```
**Determine data types**

```{r data types}
print("These are the data types for the vaccination dataset:")
sapply(vax, class)

print("These are the datatypes for the county demographics dataset:")
sapply(county, class)

```
**Identifying desired type/format for each data**

as_of_date: character -> date
vaccine_equity_metric_quartile: integer -> factor
```{r}
vax$as_of_date <- as_date(vax$as_of_date)
class(vax$as_of_date)

vax$vaccine_equity_metric_quartile <- as.factor(vax$vaccine_equity_metric_quartile)
class(vax$vaccine_equity_metric_quartile)
```

**Basic descriptives of data elements**

```{r basic descriptives}

print("Here are the simple frequencies for the county and vaccine equity metric (by quartile) variables.")

table(vax$county) #how many ZIP code time entries exist in each county
table(vax$vaccine_equity_metric_quartile) # ZIP code time entries categorized by vaccine equity metric (by quartile) 
table(county$name) #how many counties exist in the ca_county_demographic dataset


print("Here are summary statistics for numeric variables of interest.")
summary(vax$age12_plus_population)
summary(vax$persons_fully_vaccinated)
summary(vax$persons_partially_vaccinated)
summary(county$pop2012)
summary(county$med_age)
```

