---
title: 'Final Report: COVID-19 Vaccinations in California'
author: "Jessica Pak, Saira Mayet, Andrew Nguyen"
date: "12/13/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=40), tidy = T)


library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
library("sf")
library("sp")
library("spData")
library("dplyr")
library("ggplot2")
library("USAboundaries")
library("tmap")
library("shinyjs")

```

```{r import statement, echo=FALSE}
# from Milestone #2
file_path_vax <- 
  "https://data.chhs.ca.gov/dataset/ead44d40-fd63-4f9f-950a-3b0111074de8/resource/ec32eece-7474-4488-87f0-6e91cb577458/download/covid19vaccinesbyzipcode_test.csv"
file_path_county <- 
  "https://raw.githubusercontent.com/Averysaurus/reproducable_examples-/main/ca_county_demographics.csv"
vax_temp <- read.csv(file_path_vax)
county_temp <- read.csv(file_path_county)
# subset data to only include variables of interest
vax <- vax_temp %>% 
  select(-c("vaccine_equity_metric_quartile", 
            "local_health_jurisdiction", 
            "vem_source",
            "age12_plus_population",
            11:14))
county <- county_temp %>% 
  select(c("name", 
           "med_age"))
#get latest date and convert into character for use in plot captions
date <- vax %>%
  mutate(date = as.Date.character(as_of_date)) %>%
  select(- as_of_date)

current_date <- as.Date(max(date$date))
current_date_chr <- as.character(current_date)

caption <- paste("Source: California Department of Health, current to", current_date_chr)
```
## Project Problem Statement

We are monitoring COVID-19 vaccination rates among counties in California and 
in relation to age. Specifically, we are interested in whether there is any 
correlation between median age and vaccinated person prevalence at the county 
level. Utilizing two datasets, the California Census Data and the COVID-19 
Vaccine Progress Dashboard, we intend on exploring, analyzing, and visualizing
vaccination rates at the county level and the relationship between age and 
vaccination rate.

\newpage

## Methods 

Of the two datasets of interest, the first describes COVID-19 vaccination 
administration across the state of California. This is sourced from the 
California Open Data Portal for the California Department of Public Health. 
Relevant fields include date, zip code, county, and raw counts of vaccination 
status, among other population information. The dataset starts from January 5th,
2021 and is continually updated to reflect new changes in the data, most 
recently updated on November 23rd, 2021. 

The second dataset describes demographics (e.g. population, race/ethnicity, age,
household size, etc.) for each California county from 2012, which may include outdated numbers and demographics. This dataset was rehosted on Avery Richard’s GitHub, and is 
sourced from Census data.

The third dataset, or data object, was a shapefile of the state of California and its county boundaries, available in the USABoundaries R library.

## Data Cleaning & New Variable Calculations

#### Variables kept from each data source:

COVID-19 Vaccination Dashboard: 
Date, Zip Code, County, Population 5+, Number of Persons Fully Vaccinated, 
Number of Persons Partially Vaccinated, Vaccine Equity Metric Quartile 

County Census Data: 
Name, Median Age 

Geographic Shapefile Data: 
Geometry

\  

*After subsetting the data from all sources, the data were cleaned and new 
variables were created.*

#### Data Cleaning

1. Mean imputation: county-level means of fully and partially vaccinated that 
will be used to replace NA values in dataset
2. Merging relational data: county demographic dataset with vaccine 
administration dataset using key variable "county", merging county-level aggregate data with geographic shapefiles 

#### New Variables

1. Percent eligible population partially vaccinated = # of persons partially 
vaccinated / population 5+ (at county level)
2. Percent eligible population fully vaccination = # of persons fully vaccinated 
/ population 5+ (at county level)

\newpage
```{r subset latest data, echo=FALSE}
vax_latest <- vax %>%
  mutate(date = as_date(as_of_date)) %>%
  filter(date == max(date)) %>%
  select( - as_of_date)
```

```{r mean imputation,  echo=FALSE}
#aggregate county-level counts + %s for vaccination
vax_temp <- vax_latest %>%
  group_by(county) %>%
  mutate(county_partial = 
           sum(persons_partially_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm =T),
         county_fully = 
           sum(persons_fully_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm = T)) %>%
  ungroup()

#mean imputation: for ZIPs that have eligible population counts but lack vaccination numbers
vax_temp <- vax_temp %>%
  mutate(persons_partially_vaccinated_2 = 
           ifelse(is.na(persons_partially_vaccinated), 
                  age5_plus_population*county_partial,
                  persons_partially_vaccinated),
         persons_fully_vaccinated_2 = 
           ifelse(is.na(persons_fully_vaccinated),
                  age5_plus_population*county_fully, 
                  persons_fully_vaccinated))

#make second county-level aggregate with imputed data + original data
vax_temp <- vax_temp %>%
  group_by(county) %>%
  mutate(county_partial_2 = 
           sum(persons_partially_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm =T),
         county_fully_2 = 
           sum(persons_fully_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm = T),
         county_eligible_pop = 
           sum(age5_plus_population, na.rm = T)) %>%
  ungroup()

#create new dataset with only one entry per county
vax_aggregate <- vax_temp %>%
  distinct(county, .keep_all = T) %>%
  select(county, 
         county_partial_2, 
         county_fully_2, 
         county_eligible_pop) %>%
  filter(county_partial_2 != 0)
```

```{r merging datasets, echo=FALSE}
#Merging both county and vax_latest dataset
county <- county %>%
  rename(county = name)
data <- left_join(vax_aggregate, county, by="county")
```

```{r table data setup,echo=FALSE}
# final clean data
clean_data <- data %>% 
  rename(County = county,
         Partially_Vax_Rate = county_partial_2,
         Fully_Vax_Rate = county_fully_2,
         Eligible_Population = county_eligible_pop,
         Median_Age = med_age) %>% 
  mutate(Partially_Vax_Rate = round((Partially_Vax_Rate*100), digits =2),
         Fully_Vax_Rate = round((Fully_Vax_Rate*100), digits = 2),
         Eligible_Population = round(Eligible_Population, digits = 0),
         Median_Age = round(Median_Age, digits = 0)) %>% 
  arrange(County)
```


## Visualizations: Table

The Table, "Vaccination Rates for California Counties," displays the partial vaccinated rate, fully vaccinated rate, total eligible population, and median age of each county in California. 
The summarized average rates, total eligible population, and median age for the 
entire state of California is calculated and presented at the top of the table. Counties with the highest and lowest full-vaccination rates are highlighted in red.

```{r kable; table 1, echo=FALSE, warning = FALSE}
# table with descriptive statistics (averages calculated for CA as a whole) and 4 data elements
kable(clean_data, 
      booktabs = T, 
      col.names = c("County Name", 
                    "Partially Vaccinated Rate", 
                    "Fully Vaccinated Rate", 
                    "Total Eligible Population", 
                    "Median Age"), 
      align = "c", 
      caption = "Vaccination Rates for California Counties",
      format.args=list(big.mark=",")) %>% 
      add_header_above(c("California"= 1,
      "6.97"=1,
      "62.8"=1,
      "33,330,578"=1,
      "38"=1),
      bold = T) %>%
      kable_styling(latex_options = c("striped", "scale_down")) %>%
  kable_styling(font_size = 7) %>% 
  column_spec(1:5, color=if_else(clean_data$Fully_Vax_Rate == 25.67|clean_data$Fully_Vax_Rate == 82.24,"red","black","black")) %>% 
  footnote(
    general="Number of vaccinations and eligible population by zipcode are from CDPH COVID-19 Vaccination Dashboard (live dashboard), median age of each county is from California census demographic data (2012)",
    general_title="Table Data Source",
    title_format=c("italic"),
    threeparttable=T
  ) 
```
\newpage

## Visualizations: Scatterplot

The scatterplot, "Percentage of Fully Vaccinated Californians by Median Age, 2021," shows the rate of fully vaccinated individuals stratified by median age of California residents, as measured from California Department of Public Health’s latest online data repository for COVID-19. A trend line was fitted to follow the points on the plot, and a general trend of increasing fully vaccinated rates can be seen near the median age of 36-38.

```{r age vs. full vaccination rate plot, echo=FALSE, warning = FALSE}
# Scatterplot of Median Age vs Fully Vaccinated Rate

ggplot(data = clean_data, aes(x= Median_Age,y= Fully_Vax_Rate)) +
  geom_point() +
  geom_smooth(col = "navy blue", size = 0.5) + 
  scale_y_continuous(limits=c(0, 100)) +
  labs(x= "Median age", y= "Rate(s) of full vaccination", title= "Percentage of Fully Vaccinated Californians by Median Age, 2021", caption=caption) +
  theme_classic()
```
\newpage 

## Visualizations: Maps  

The map shows the rates of full vaccination, by county, symbolized by a color ramp (from low to high rates of vaccination going from yellow to dark red). A color ramp was made with classes of 10 percent each. A second map shows the median age of each country, symbolized by a purple color ramp (younger to older median age going from light purple to darker purple). The five-step color ramp ranges from 30 to 55, with each step covering five years of age. 

```{r Map, echo=FALSE, warning = FALSE}

#get CA boundary with high definition
#ca <- USAboundaries::us_states(resolution = "high", states = "CA")

#CA county boundary
ca_co <- USAboundaries::us_counties(resolution = "high", states = "CA") %>%
  select(- state_name)

#Join dataset with sf polygon
clean_data_rename <- clean_data %>%
  rename(name = County)

ca_covid <- dplyr::left_join(clean_data_rename, ca_co, by = "name")

#Make joined dataset into sf object
ca_covid <- st_as_sf(ca_covid, sf_column_name = "geometry")

#Create first map
legend_title = " "
tm_shape(ca_covid) + 
  tm_fill(col = "Fully_Vax_Rate", 
          title = legend_title,
          labels = c('20 - 30%','30 - 40%','40 - 50%','50 - 60%', '60 - 70%', '70 - 80%', '80 - 90%')) +
  tm_compass(type = "8star", 
             position = c("left", "bottom"),
             size = 2) +
  tm_layout(scale = 1, 
            title = "Percent of population fully vaccinated in California, by county",
            frame.lwd = 0,
            inner.margins = 0.1,
            legend.position = c("right", "top")) +
  tm_credits(caption, size = 0.5)

#Create second map
tm_shape(ca_covid) + 
  tm_fill(col = "Median_Age", 
          title = legend_title,
          palette = "Purples") +
  tm_compass(type = "8star", 
             position = c("left", "bottom"),
             size = 2) +
  tm_layout(scale = 1, 
            title = "Median age, by county",
            frame.lwd = 0,
            inner.margins = 0.1,
            legend.position = c("right", "top")) +
  tm_credits("Source: California demographic census data from 2012", size = 0.5)
```

\newpage 

## Visualizations: Longitudinal Plot  

This last visualization features points plotted by month and percent of full vaccination in aggregate ZIP codes falling under four categories of the Vaccine Equity Metric (VEM) quartiles, which is based on demographic indicators of socioeconomic disadvantage and health vulnerability (where 1 is least disadvantaged/vulnerable and 4 and most disadvantaged/vulnerable). 
```{r longitudinal, echo=FALSE, warning = FALSE}
#get data per day at ZIP code level
vax_longitudinal <- read.csv(file_path_vax) %>%
  group_by(county) %>%
  mutate(county_partial = 
           sum(persons_partially_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm =T),
         county_fully = 
           sum(persons_fully_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm = T)) %>%
  ungroup() %>%
  mutate(persons_partially_vaccinated_2 = 
           ifelse(is.na(persons_partially_vaccinated), 
                  age5_plus_population*county_partial,
                  persons_partially_vaccinated),
         persons_fully_vaccinated_2 = 
           ifelse(is.na(persons_fully_vaccinated),
                  age5_plus_population*county_fully, 
                  persons_fully_vaccinated)) %>%
   group_by(county) %>%
  mutate(county_partial_2 = 
           sum(persons_partially_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm =T),
         county_fully_2 = 
           sum(persons_fully_vaccinated, na.rm = T)/
           sum(age5_plus_population, na.rm = T),
         county_eligible_pop = 
           sum(age5_plus_population, na.rm = T)) %>%
  ungroup()

#create aggregate dataset at VEM quartile level
vax_long_agg <- vax_longitudinal %>%
  group_by(as_of_date, vaccine_equity_metric_quartile) %>%
  summarize(percent_full_vem = 
              sum(persons_fully_vaccinated, na.rm = T)/
              sum(age5_plus_population, na.rm = T)) %>%
  ungroup() %>%
  filter(is.na(vaccine_equity_metric_quartile) == FALSE) %>%
  mutate(date = as.Date.character(as_of_date)) %>%
  select(- as_of_date)

#Create longitudinal scatterplot visualization
ggplot(data = vax_long_agg, aes(x = date, y = percent_full_vem)) +
  geom_point(stat = "identity", aes(col = vaccine_equity_metric_quartile)) +
  theme_classic() +
  labs(x = "Date", 
       y = "Percent fully vaccinated", 
       title = "Rates of full vaccination across Vaccine Equity Metric Quartiles", caption = caption) +
  guides(color=guide_legend(title="Vaccine Equity Metric Quartiles")) +
  scale_color_continuous(labels = 
                           c("1 - Most Disadvantaged", "2", "3", "4 - Least Disadvantaged")) +
  scale_y_continuous(labels = scales::percent)

```

\newpage

## Discussion

Our analyses and data visualizations show that the highest vaccination rates 
were seen among the age group between 35-40. The general trend seen for this age
group exhibited a full vaccination rate of approximately 70-73%, with rates seen 
above 75% as well. Outliers were seen with higher rates of vaccination, approximately 85% for a 
median age of 44. Outliers were also found for lower rates of vaccination of 
approximately 25% for the median age of 37. According to our results, the minimum median age of persons receiving COVID-19 Vaccination in the state of California is 30, for Merced county. The maximum median age was found to be 51 in Sierra county. 

Overall, in acknowledgement to our original problem statement, we were not able to conclude that there seems to be a strong linear relationship between median age and rates of full vaccination across California counties. Further statistical analysis would be needed to establish a robust correlative relationship between the two variables.

However, we explored other associations that may be interesting to follow up in future studies, including geography of vaccination rates in counties such as San Francisco, Santa Clara, and Marin county with high full vaccination rates at or above 80% and median ages of 38, 36, 44, respectively. In addition, it can be seen that rates of vaccination follow, from highest to lowest, the order of vulnerability derived from the Vaccine Equity Metric (VEM), from least to most -- suggesting areas of improvement for equitable vaccine delivery. The aggregate of the least disadvantaged areas (VEM group 4) sees vaccination rates of  77%, whereas the most disadvantaged areas (VEM group 1) trails behind at 59%. 

\newpage

## Bonus discussion
**Challenges**

***Understanding and addressing confusing data***

Some ZIPs had abnormally high vaccination totals compared to their population totals (sometimes having rates higher than 100%); we had to better understand where this data anomaly came from (e.g. some people chose to vaccinate outside of their home ZIP, therefore seemingly inflating numbers) and how to address it. In the end, no observations were removed due to this issue. rather, we aggregated vaccination numbers at the county level, acknowledging that the vaccinated population is still captured accurately at a larger geography in comparison to the total population.
    
***Working with a live dataset***

Changes in column names, vaccination eligible age, and temporality of data provided an extra layer of consideration when we were devising code that could be replicable and current to the updating dataset without maintenance. We addressed this by creating variables that returned the latest date using the _lubridate_ package to ensure the results and visualizations would be up-to-date every time the .rmd was ran without manual user intervention.
    
**Something new we learned**

***Making maps with R*** 

The process of making maps with R is greatly simplified with the ability to treat geographic objects (shapefiles) as dataframes and/or dplyr-compatible tibbles. This meant that county-level data was easily appended to geographic geometry. The process of visualizing this data drew heavily upon the coursework in visualization (e.g. ggplot, table creation) and was intuitive.
    
***Making HTML slides with R***

Creation of visually-appealing and easy to navigate slidedecks through .rmd files with the slidy-presentation preset.
    

